---
title: "nov_20"
author: "xiaowen xu"
format: gfm
editor: visual
---

Pertussis (more commonly known as whooping cough) is a highly contagious respiratory disease caused by the bacterium Bordetella pertussis

The United States Centers for Disease Control and Prevention (CDC) has been compiling reported pertussis case numbers since 1922 in their National Notifiable Diseases Surveillance System (NNDSS). We can view this data on the CDC website \[here\]:(https://www.cdc.gov/pertussis/php/surveillance/pertussis-cases-by-year.html?CDC_AAref_Val=https://www.cdc.gov/pertussis/surv-reporting/cases-by-year.html)

Q1. With the help of the R “addin” package datapasta assign the CDC pertussis case number data to a data frame called cdc and use ggplot to make a plot of cases numbers over time.

we need to scrape this data so we can do stuff with it in R. let's try the datapasta package

```{r}

cdc <- data.frame(
  year = c(
    1922L, 1923L, 1924L, 1925L, 1926L,
    1927L, 1928L, 1929L, 1930L, 1931L,
    1932L, 1933L, 1934L, 1935L, 1936L, 1937L,
    1938L, 1939L, 1940L, 1941L, 1942L,
    1943L, 1944L, 1945L, 1946L, 1947L, 1948L,
    1949L, 1950L, 1951L, 1952L, 1953L, 1954L,
    1955L, 1956L, 1957L, 1958L, 1959L,
    1960L, 1961L, 1962L, 1963L, 1964L, 1965L,
    1966L, 1967L, 1968L, 1969L, 1970L,
    1971L, 1972L, 1973L, 1974L, 1975L, 1976L,
    1977L, 1978L, 1979L, 1980L, 1981L,
    1982L, 1983L, 1984L, 1985L, 1986L, 1987L,
    1988L, 1989L, 1990L, 1991L, 1992L, 1993L,
    1994L, 1995L, 1996L, 1997L, 1998L,
    1999L, 2000L, 2001L, 2002L, 2003L, 2004L,
    2005L, 2006L, 2007L, 2008L, 2009L,
    2010L, 2011L, 2012L, 2013L, 2014L, 2015L,
    2016L, 2017L, 2018L, 2019L, 2020L,
    2021L, 2022L,2024L
  ),
  cases = c(
    107473, 164191, 165418, 152003,
    202210, 181411, 161799, 197371, 166914,
    172559, 215343, 179135, 265269, 180518,
    147237, 214652, 227319, 103188, 183866,
    222202, 191383, 191890, 109873, 133792,
    109860, 156517, 74715, 69479, 120718,
    68687, 45030, 37129, 60886, 62786,
    31732, 28295, 32148, 40005, 14809, 11468,
    17749, 17135, 13005, 6799, 7717, 9718,
    4810, 3285, 4249, 3036, 3287, 1759,
    2402, 1738, 1010, 2177, 2063, 1623, 1730,
    1248, 1895, 2463, 2276, 3589, 4195,
    2823, 3450, 4157, 4570, 2719, 4083, 6586,
    4617, 5137, 7796, 6564, 7405, 7298,
    7867, 7580, 9771, 11647, 25827, 25616,
    15632, 10454, 13278, 16858, 27550, 18719,
    48277, 28639, 32971, 20762, 17972,
    18975, 15609, 18617, 6124, 2116, 3044,23544
  )
)

cdc
```

```{r}
library(ggplot2) 
baseplot <- ggplot(cdc) +
  aes(year, cases) +
  geom_point() +
  geom_line() + 
  labs(title = "Yearly Cases", x = "Year", y = "Number of Cases")
baseplot

```

let's add the date of wP, and the switch to new aP vaccine(1996)

```{r}
baseplot+
  theme_bw()+
  geom_vline(xintercept=1946,col="blue")+
  geom_vline(xintercept = 1996, col = "red")+
  geom_vline(xintercept=2020,col="gray")+
  geom_vline(xintercept=2004,col="black")
```

##CMI-PB(computational models of immunity-Pertussis Boost)

this projects collects and makes freely available data about the immune response to pertussis vaccination

you can access this data via an API which returns JSON format (key:value pairs).

you cna use the jsonlite package and its read_json function

```{r}
library(jsonlite)

subject <- read_json("https://www.cmi-pb.org/api/v5/subject",simplifyVector = TRUE)
```

kets have a wee peak and explore of this

```{r}
head(subject)
```

Q how many subjects do we have?

```{r}
nrow(subject)
```

Q how many male/female do we have?

```{r}
table(subject$biological_sex)
```

how many wP and aP do we have?

```{r}
table(subject$infancy_vac)
```

break down of biological sex and race?

```{r}
table(subject$biological_sex, subject$race)

```

does this break down reflect US population,no

```{r}
table(subject$dataset)
```

```{r}
library(jsonlite)

specimen <- read_json("https://www.cmi-pb.org/api/v5/specimen",simplifyVector = TRUE)
```

```{r}

library(jsonlite)

ab_titer <- read_json("https://www.cmi-pb.org/api/v5/plasma_ab_titer",simplifyVector = TRUE)
```

```{r}
head(specimen)
```

```{r}
head(ab_titer)
```

we want to merge or join these tables so we can have all the info we need about a give antibody measurement

```{r}
library(dplyr)

meta <- inner_join(subject, specimen)
```

```{r}
head(meta)
```

and add one last join of ab_titer and meta

```{r}
library(dplyr)

abdata <- inner_join(ab_titer, meta)
```

```{r}
head(abdata)
```

```{r}
nrow(abdata)
```

```{r}
head(abdata)
```

```{r}
table(abdata$isotype)
```

```{r}
table(abdata$antigen)
```

lets begin with IgG

```{r}
igg <-filter(abdata,isotype=="IgG")
head(igg)
```

make a boxplot of IgG antigen levels, this will be MFI_normalised versus antigen

```{r}
library(ggplot2)

ggplot(abdata) +
  aes(x = MFI_normalised, y = antigen) + 
  geom_boxplot() + 
  labs(
    title = "Boxplot of IgG Antigen Levels",
    x = "MFI_normalized",
    y = "antigen"
  )

```

```{r}
library(ggplot2)

ggplot(abdata) +
  aes(x = MFI_normalised, y = antigen,col=infancy_vac) + 
  geom_boxplot() + 
  labs(
    title = "Boxplot of IgG Antigen Levels",
    x = "MFI_normalized",
    y = "antigen"
  )

```

ideally I would like to see how these ab levels change over time relative to the booster shot

```{r}
table(abdata$visit)
```

```{r}
ggplot(abdata) +
  aes(x = MFI_normalised, y = antigen) + 
  geom_boxplot() + 
  labs(
    title = "Boxplot of IgG Antigen Levels",
    x = "MFI_normalized",
    y = "antigen"
  )+
  facet_wrap(~infancy_vac)
```

```{r}

ggplot(abdata) +
  aes(x = MFI_normalised, y = antigen,col=infancy_vac) + 
  geom_boxplot() + 
  labs(
    title = "Boxplot of IgG Antigen Levels",
    x = "MFI_normalized",
    y = "antigen"
  )+
  facet_wrap(~visit)

```

```{r}
ggplot(abdata) +
  aes(x = MFI_normalised, y = antigen,col=infancy_vac) + 
  geom_boxplot() + 
  labs(
    title = "Boxplot of IgG Antigen Levels",
    x = "MFI_normalized",
    y = "antigen"
  )+
  facet_wrap(~visit)
```

```{r}
filter(igg, antigen=="PT",dataset=="2021_dataset") %>%
  ggplot() +
  aes(MFI_normalised, col=infancy_vac) +
  geom_boxplot(show.legend = FALSE) +
  facet_wrap(vars(visit)) +
  theme_bw()

```

```{r}
abdata.21 <- abdata %>% filter(dataset == "2021_dataset")

library(dplyr)
library(ggplot2)

average_data <- abdata.21 %>%
  filter(isotype == "IgG", antigen == "PT") %>%
  group_by(infancy_vac, planned_day_relative_to_boost) %>%
  summarise(avg_MFI_normalised = mean(MFI_normalised, na.rm = TRUE), .groups = "drop")

abdata.21 %>%
  filter(isotype == "IgG", antigen == "PT") %>%
  ggplot(aes(x = planned_day_relative_to_boost, y = MFI_normalised, col = infancy_vac, group = subject_id)) +
  geom_point() +
  geom_line() +
  # Average curves
  geom_line(
    data = average_data, 
    aes(x = planned_day_relative_to_boost, y = avg_MFI_normalised, col = infancy_vac), 
    size = 1.5, inherit.aes = FALSE
  ) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_vline(xintercept = 14, linetype = "dashed") +
  labs(
    title = "2021 dataset IgG PT",
    subtitle = "Dashed lines indicate day 0 (pre-boost) and 14 (apparent peak levels)",
    x = "Planned Day Relative to Boost",
    y = "MFI Normalised"
  ) +
  theme_minimal()


```
